Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-04-01 manu-draft.qmd
data summary table added, more revisions

load(file = here('data/speidat.RData'))
load(file = here('data/gamfit.RData'))
load(file = here('tabs/temptrndtab.RData'))
load(file = here('tabs/saltrndtab.RData'))
load(file = here('tabs/mixdaytab.RData'))
load(file = here('tabs/modtab.RData'))
load(file = here('data/sgmods.RData'))
```

@@

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created from aerial images collected specifically to map seagrass and are acquired during a flight window from December and February. The maps are created by photointerpretation of image signatures coupled with a robust field verification and accuracy assessment. The maps provide a spatial estimate of seagrass cover at the landscape scale and is irrespective of species. Complementary to the SWFWMD seagrass mapping program is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on cover-abundance, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at fixed stations in the major bay segments since the early 1970s (Figure [-@fig-map]b). The 45 stations with the longest and most complete temporal record from 1975 to present were used herein. Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006], although 1975 bottom salinity used middle water column sampling since the former was not available until the following year. Most samples are collected from mid-morning to early afternoon. Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of water quality change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers. For comparison to water quality conditions, daily air temperature (Tampa International Airport \[TIA\] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the most consistent period of record covered by the water quality samples (i.e., 1975 to present for the EPC data). The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data. Regional precipitation summaries were obtained directly from the SWFWMD (https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region). Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. All climate data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment.

@@

Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs. GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with time period and a main effect for bay segment were also included as above. GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. The seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrasses. Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.

@@

>>>>

load(file = here('data/speidat.RData'))
load(file = here('data/gamfit.RData'))
load(file = here('tabs/dattab.RData'))
load(file = here('tabs/temptrndtab.RData'))
load(file = here('tabs/saltrndtab.RData'))
load(file = here('tabs/mixdaytab.RData'))
load(file = here('tabs/modtab.RData'))
load(file = here('data/sgmods.RData'))
```

@@

Two primary sources of data have been used to track seagrass change over time in Tampa Bay (@tbl-dattab). The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created from aerial images collected specifically to map seagrass and are acquired during a flight window from December and February. The maps are created by photointerpretation of image signatures coupled with a robust field verification and accuracy assessment. The maps provide a spatial estimate of seagrass cover at the landscape scale and is irrespective of species. Complementary to the SWFWMD seagrass mapping program is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on cover-abundance, frequency occurrence [number of sample points with seagrass divided by total points on a transect, as in @sherwood2017], and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year. The transect data are collected each year, allowing inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay (@tbl-dattab). These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at fixed stations in the major bay segments since the early 1970s (Figure [-@fig-map]b). The 45 stations with the longest and most complete temporal record from 1975 to present were used herein. Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006], although 1975 bottom salinity used middle water column sampling since the former was not available until the following year. Most samples are collected from mid-morning to early afternoon. Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of water quality change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change drivers. For comparison to water quality conditions, daily air temperature (Tampa International Airport [TIA] National Weather Service site) and precipitation (SWFWMD area-weighted watershed summaries) were used to characterize regional conditions for the most consistent period of record covered by the water quality samples (i.e., 1975 to present for the EPC data, @tbl-dattab). The rnoaa R package [@chamberlain23] was used to obtain the TIA temperature data. Regional precipitation summaries were obtained directly from the SWFWMD (https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region). Only rainfall data for the wet season (June to September) were evaluated for trends, whereas the complete record was used to calculate the Standardized Precipitation Index [SPI, @begueria2013] to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. All climate data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment.

@@

Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with time period and a main effect for bay segment were also included as above. GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. The seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrasses. Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.

@@

```{r}
#| label: tbl-dattab
#| tbl-cap: "Summary of datasets used for trend analyses and comparisons with seagrass datasets.  Temporal and spatial scales are those used for analysis.  SWFWMD: Southwest Florida Water Management District."
dattab
```
